shorts: An R Package for Modeling Short Sprints

### 2021-04-14 paper.tex
Reduced version for the PeerJ

\usepackage{flafter}

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 s) of maximal efforts (\protect\hyperlink{ref-mangineSpeedForcePower2014}{Mangine et al. 2014}), whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) (\protect\hyperlink{ref-ward-smithEnergyConversionStrategies2001}{Ward-Smith 2001}), but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)(\protect\hyperlink{ref-brownAssessmentLinearSprinting2004}{Brown, Vescovi, and Vanheest 2004}). Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). Radar and laser technology are frequently used laboratory-grade methods (\protect\hyperlink{ref-buchheitMechanicalDeterminantsAcceleration2014}{Buchheit et al. 2014}; \protect\hyperlink{ref-edwardsSprintAccelerationCharacteristics2020}{Edwards et al. 2020}; \protect\hyperlink{ref-jimenez-reyesRelationshipVerticalHorizontal2018}{Jiménez-Reyes et al. 2018}; \protect\hyperlink{ref-marcote-pequenoAssociationForceVelocity2019}{Marcote-Pequeño et al. 2019}) but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \textbf{shorts} package (\protect\hyperlink{ref-R-shorts}{Jovanovic 2020}), written in the R language (\protect\hyperlink{ref-R-base}{R Core Team 2020}), represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}).

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/four-athletes-kinematics-1}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/four-athletes-profile-1}

@@

Besides providing \emph{residual standard error} (RSE), \textbf{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the \texttt{nls()} function:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{summary}\NormalTok{(m1)}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Formula: corrected\_time \textasciitilde{} TAU * I(LambertW::W({-}exp(1)\^{}({-}distance/(MSS * }
\CommentTok{\#\textgreater{}     TAU) {-} 1))) + distance/MSS + TAU}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Parameters:}
\CommentTok{\#\textgreater{}     Estimate Std. Error t value Pr(\textgreater{}|t|)    }
\CommentTok{\#\textgreater{} MSS   9.0121     0.0155     581 0.000003 ***}
\CommentTok{\#\textgreater{} TAU   1.3083     0.0066     198 0.000025 ***}
\CommentTok{\#\textgreater{} {-}{-}{-}}
\CommentTok{\#\textgreater{} Signif. codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Residual standard error: 0.00249 on 2 degrees of freedom}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Number of iterations to convergence: 4 }
\CommentTok{\#\textgreater{} Achieved convergence tolerance: 0.00000311}
\end{Highlighting}
\end{Shaded}

To extract estimated model parameters, use S3 \texttt{coef()} method:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{coef}\NormalTok{(m1)}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}                9.01                1.31                6.89 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}               15.52                0.00                0.00}
\end{Highlighting}
\end{Shaded}

@@

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-4-1} \end{center}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Predict time at distance}
\FunctionTok{predict\_time\_at\_distance}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 2.17 3.43 4.60 5.73}

\CommentTok{\# Predict acceleration at time}
\FunctionTok{predict\_acceleration\_at\_time}\NormalTok{(}
  \AttributeTok{time =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{6}\NormalTok{),}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 6.8884 3.2075 1.4935 0.6954 0.3238 0.1508 0.0702}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# To calculate horizontal force produced}
\FunctionTok{predict\_force\_at\_distance}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU,}
  \CommentTok{\# Additional parameters forwarded to get\_air\_resistance}
  \CommentTok{\# Otherwise, defaults are used}
  \AttributeTok{bodymass =} \DecValTok{80}\NormalTok{,}
  \AttributeTok{bodyheight =} \FloatTok{1.85}\NormalTok{,}
  \AttributeTok{barometric\_pressure =} \DecValTok{780}\NormalTok{,}
  \AttributeTok{air\_temperature =} \DecValTok{20}\NormalTok{,}
  \AttributeTok{wind\_velocity =} \FloatTok{0.5}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 119.0  58.6  36.9  28.2}

\CommentTok{\# To calculate power produced}
\FunctionTok{predict\_power\_at\_distance}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU,}
  \CommentTok{\# Additional parameters forwarded to get\_air\_resistance}
  \CommentTok{\# Otherwise, defaults are used}
  \AttributeTok{bodymass =} \DecValTok{80}\NormalTok{,}
  \AttributeTok{bodyheight =} \FloatTok{1.85}\NormalTok{,}
  \AttributeTok{barometric\_pressure =} \DecValTok{780}\NormalTok{,}
  \AttributeTok{air\_temperature =} \DecValTok{20}\NormalTok{,}
  \AttributeTok{wind\_velocity =} \FloatTok{0.5}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 868 490 323 251}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df }\OtherTok{\textless{}{-}} \FunctionTok{predict\_kinematics}\NormalTok{(}
\NormalTok{  m1,}
  \AttributeTok{max\_time =} \DecValTok{6}\NormalTok{,}
  \AttributeTok{frequency =} \DecValTok{100}\NormalTok{,}
  \CommentTok{\# Additional parameters forwarded to get\_air\_resistance}
  \CommentTok{\# Otherwise, defaults are used}
  \AttributeTok{bodymass =} \DecValTok{80}\NormalTok{,}
  \AttributeTok{bodyheight =} \FloatTok{1.85}\NormalTok{,}
  \AttributeTok{barometric\_pressure =} \DecValTok{780}\NormalTok{,}
  \AttributeTok{air\_temperature =} \DecValTok{20}\NormalTok{,}
  \AttributeTok{wind\_velocity =} \FloatTok{0.5}
\NormalTok{)}

\FunctionTok{head}\NormalTok{(df)}
\CommentTok{\#\textgreater{}   time distance velocity acceleration bodymass}
\CommentTok{\#\textgreater{} 1 0.00 0.000000   0.0000         6.89       80}
\CommentTok{\#\textgreater{} 2 0.01 0.000344   0.0686         6.84       80}
\CommentTok{\#\textgreater{} 3 0.02 0.001371   0.1367         6.78       80}
\CommentTok{\#\textgreater{} 4 0.03 0.003076   0.2043         6.73       80}
\CommentTok{\#\textgreater{} 5 0.04 0.005455   0.2714         6.68       80}
\CommentTok{\#\textgreater{} 6 0.05 0.008502   0.3379         6.63       80}
\CommentTok{\#\textgreater{}   net\_horizontal\_force air\_resistance horizontal\_force}
\CommentTok{\#\textgreater{} 1                  551        0.07536              551}
\CommentTok{\#\textgreater{} 2                  547        0.05609              547}
\CommentTok{\#\textgreater{} 3                  543        0.03978              543}
\CommentTok{\#\textgreater{} 4                  539        0.02636              539}
\CommentTok{\#\textgreater{} 5                  534        0.01576              534}
\CommentTok{\#\textgreater{} 6                  530        0.00792              530}
\CommentTok{\#\textgreater{}   horizontal\_force\_relative vertical\_force resultant\_force}
\CommentTok{\#\textgreater{} 1                      6.89            785             959}
\CommentTok{\#\textgreater{} 2                      6.84            785             957}
\CommentTok{\#\textgreater{} 3                      6.78            785             954}
\CommentTok{\#\textgreater{} 4                      6.73            785             952}
\CommentTok{\#\textgreater{} 5                      6.68            785             950}
\CommentTok{\#\textgreater{} 6                      6.63            785             947}
\CommentTok{\#\textgreater{}   resultant\_force\_relative power relative\_power    RF}
\CommentTok{\#\textgreater{} 1                     12.0   0.0          0.000 0.575}
\CommentTok{\#\textgreater{} 2                     12.0  37.5          0.469 0.572}
\CommentTok{\#\textgreater{} 3                     11.9  74.2          0.928 0.569}
\CommentTok{\#\textgreater{} 4                     11.9 110.0          1.375 0.566}
\CommentTok{\#\textgreater{} 5                     11.9 145.0          1.813 0.563}
\CommentTok{\#\textgreater{} 6                     11.8 179.2          2.241 0.560}
\CommentTok{\#\textgreater{}   force\_angle}
\CommentTok{\#\textgreater{} 1        54.9}
\CommentTok{\#\textgreater{} 2        55.1}
\CommentTok{\#\textgreater{} 3        55.3}
\CommentTok{\#\textgreater{} 4        55.5}
\CommentTok{\#\textgreater{} 5        55.7}
\CommentTok{\#\textgreater{} 6        55.9}
\end{Highlighting}
\end{Shaded}

@@

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-9-1} \end{center}

@@

Additional information about mixed-effects model performed using the \texttt{nlme} package (\protect\hyperlink{ref-R-nlme}{Pinheiro, Bates, and R-core 2020}) can be obtained using \texttt{summary()} function:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{summary}\NormalTok{(m2)}
\CommentTok{\#\textgreater{} Nonlinear mixed{-}effects model fit by maximum likelihood}
\CommentTok{\#\textgreater{}   Model: corrected\_time \textasciitilde{} TAU * I(LambertW::W({-}exp(1)\^{}({-}distance/(MSS *      TAU) {-} 1))) + distance/MSS + TAU }
\CommentTok{\#\textgreater{}   Data: train }
\CommentTok{\#\textgreater{}     AIC   BIC logLik}
\CommentTok{\#\textgreater{}   {-}75.1 {-}66.7   43.5}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Random effects:}
\CommentTok{\#\textgreater{}  Formula: list(MSS \textasciitilde{} 1, TAU \textasciitilde{} 1)}
\CommentTok{\#\textgreater{}  Level: athlete}
\CommentTok{\#\textgreater{}  Structure: General positive{-}definite, Log{-}Cholesky parametrization}
\CommentTok{\#\textgreater{}          StdDev Corr }
\CommentTok{\#\textgreater{} MSS      1.066  MSS  }
\CommentTok{\#\textgreater{} TAU      0.178  0.877}
\CommentTok{\#\textgreater{} Residual 0.026       }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Fixed effects:  MSS + TAU \textasciitilde{} 1 }
\CommentTok{\#\textgreater{}     Value Std.Error DF t{-}value p{-}value}
\CommentTok{\#\textgreater{} MSS  8.06     0.495 24   16.30       0}
\CommentTok{\#\textgreater{} TAU  0.66     0.084 24    7.82       0}
\CommentTok{\#\textgreater{}  Correlation: }
\CommentTok{\#\textgreater{}     MSS  }
\CommentTok{\#\textgreater{} TAU 0.874}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Standardized Within{-}Group Residuals:}
\CommentTok{\#\textgreater{}    Min     Q1    Med     Q3    Max }
\CommentTok{\#\textgreater{} {-}1.909 {-}0.605  0.154  0.523  1.129 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Number of Observations: 30}
\CommentTok{\#\textgreater{} Number of Groups: 5}
\end{Highlighting}
\end{Shaded}

S3 method \texttt{coef()} when applied on mixed-model result will return both the fixed and random effects:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{coef}\NormalTok{(m2)}
\CommentTok{\#\textgreater{} $fixed}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}               8.065               0.655              12.309 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}              24.818               0.000               0.000 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} $random}
\CommentTok{\#\textgreater{}     athlete  MSS   TAU  MAC PMAX time\_correction}
\CommentTok{\#\textgreater{} 1     James 9.69 0.847 11.4 27.7               0}
\CommentTok{\#\textgreater{} 2       Jim 7.83 0.505 15.5 30.4               0}
\CommentTok{\#\textgreater{} 3      John 7.78 0.727 10.7 20.8               0}
\CommentTok{\#\textgreater{} 4 Kimberley 8.57 0.802 10.7 22.9               0}
\CommentTok{\#\textgreater{} 5  Samantha 6.45 0.395 16.3 26.4               0}
\CommentTok{\#\textgreater{}   distance\_correction}
\CommentTok{\#\textgreater{} 1                   0}
\CommentTok{\#\textgreater{} 2                   0}
\CommentTok{\#\textgreater{} 3                   0}
\CommentTok{\#\textgreater{} 4                   0}
\CommentTok{\#\textgreater{} 5                   0}
\end{Highlighting}
\end{Shaded}

To create a simple plot of the model, use S3 \texttt{plot()} method:

@@

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-14-1} \end{center}

@@

\FunctionTok{ggplot}\NormalTok{(}
  \FunctionTok{filter}\NormalTok{(df, distance }\SpecialCharTok{\textless{}} \DecValTok{40}\NormalTok{),}
  \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ distance, }\AttributeTok{y =}\NormalTok{ value, }\AttributeTok{group =}\NormalTok{ athlete, }\AttributeTok{color =}\NormalTok{ athlete)}
\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{(}\DecValTok{8}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{name, }\AttributeTok{scales =} \StringTok{"free\_y"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{geom\_line}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.7}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{ylab}\NormalTok{(}\ConstantTok{NULL}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{xlab}\NormalTok{(}\StringTok{"Distance (m)"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}
    \AttributeTok{legend.position =} \StringTok{"top"}\NormalTok{,}
    \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-15-1} \end{center}

@@

\NormalTok{m3 }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_radar}\NormalTok{(}
  \AttributeTok{velocity =}\NormalTok{ sprint\_velocity,}
  \AttributeTok{time =}\NormalTok{ sprint\_time}
\NormalTok{)}

\NormalTok{m3}
\CommentTok{\#\textgreater{} Estimated model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}                9.00                1.30                6.92 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}               15.58                0.00                0.00 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Model fit estimators}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}       RSE R\_squared    minErr    maxErr maxAbsErr      RMSE }
\CommentTok{\#\textgreater{}   0.00327   1.00000  {-}0.00406   0.00532   0.00532   0.00276 }
\CommentTok{\#\textgreater{}       MAE      MAPE }
\CommentTok{\#\textgreater{}   0.00207       NaN}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(m3) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{(}\DecValTok{8}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-17-1} \end{center}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m3\_weighted }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_radar}\NormalTok{(}
  \AttributeTok{velocity =}\NormalTok{ sprint\_velocity,}
  \AttributeTok{time =}\NormalTok{ sprint\_time,}
  \AttributeTok{weights =} \DecValTok{1} \SpecialCharTok{/}\NormalTok{ (sprint\_velocity }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}
\NormalTok{)}

\NormalTok{m3\_weighted}
\CommentTok{\#\textgreater{} Estimated model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}                9.00                1.30                6.92 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}               15.58                0.00                0.00 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Model fit estimators}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}       RSE R\_squared    minErr    maxErr maxAbsErr      RMSE }
\CommentTok{\#\textgreater{}   0.00108   1.00000  {-}0.00406   0.00534   0.00534   0.00276 }
\CommentTok{\#\textgreater{}       MAE      MAPE }
\CommentTok{\#\textgreater{}   0.00206       NaN}
\end{Highlighting}
\end{Shaded}

>>>>

\usepackage{flafter}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 s) of maximal efforts (\protect\hyperlink{ref-mangineSpeedForcePower2014}{Mangine et al. 2014}), whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) (\protect\hyperlink{ref-ward-smithEnergyConversionStrategies2001}{Ward-Smith 2001}), but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)(\protect\hyperlink{ref-brownAssessmentLinearSprinting2004}{Brown, Vescovi, and Vanheest 2004}). Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). Radar and laser technology are frequently used laboratory-grade methods (\protect\hyperlink{ref-buchheitMechanicalDeterminantsAcceleration2014}{Buchheit et al. 2014}; \protect\hyperlink{ref-edwardsSprintAccelerationCharacteristics2020}{Edwards et al. 2020}; \protect\hyperlink{ref-jimenez-reyesRelationshipVerticalHorizontal2018}{Jiménez-Reyes et al. 2018}; \protect\hyperlink{ref-marcote-pequenoAssociationForceVelocity2019}{Marcote-Pequeño et al. 2019}) but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance.

The \textbf{shorts} package (\protect\hyperlink{ref-R-shorts}{Jovanovic 2020}), written in the R language (\protect\hyperlink{ref-R-base}{R Core Team 2020}), represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). To best of our knowledge, scientist, researchers, and coaches have been performing short sprints modeling using the built-in solver function of Excel (Microsoft Corporation, Redmond, Washington, United States) (\protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2017}{J. B. Morin 2017}; \protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2019}{Jean-Benoit Morin and Samozino 2019}; \protect\hyperlink{ref-stenrothForcevelocityProfilingIce2020}{Stenroth, Vartiainen, and Karjalainen 2020}; \protect\hyperlink{ref-stenrothSpreadsheetSprintAcceleration2020}{Stenroth and Vartiainen 2020}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}; \protect\hyperlink{ref-clarkNFLCombine40Yard2017}{Clark et al. 2017}; \protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}), which makes the \textbf{shorts} package major improvement in ease-of-use, speed, transparency, reproducibility, and more feature-rich model fitting.

@@

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/four-athletes-kinematics-1}

@@

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/four-athletes-profile-1}

@@

Besides providing \emph{residual standard error} (RSE), \textbf{shorts} functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the \texttt{nls()} function (i.e., by using the S3 \texttt{summary()} method). To extract estimated model parameters, use S3 \texttt{coef()} method.

@@

\begin{center}\includegraphics[width=0.9\linewidth]{paper_files/figure-latex/unnamed-chunk-2-1} \end{center}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Predict time at distance}
\FunctionTok{predict\_time\_at\_distance}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 2.17 3.43 4.60 5.73}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# To calculate horizontal force produced}
\FunctionTok{predict\_force\_at\_distance}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_distance,}
  \AttributeTok{MSS =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{MSS,}
  \AttributeTok{TAU =}\NormalTok{ m1}\SpecialCharTok{$}\NormalTok{parameters}\SpecialCharTok{$}\NormalTok{TAU,}
  \CommentTok{\# Additional parameters forwarded to get\_air\_resistance}
  \CommentTok{\# Otherwise, defaults are used}
  \AttributeTok{bodymass =} \DecValTok{80}\NormalTok{,}
  \AttributeTok{bodyheight =} \FloatTok{1.85}\NormalTok{,}
  \AttributeTok{barometric\_pressure =} \DecValTok{780}\NormalTok{,}
  \AttributeTok{air\_temperature =} \DecValTok{20}\NormalTok{,}
  \AttributeTok{wind\_velocity =} \FloatTok{0.5}
\NormalTok{)}
\CommentTok{\#\textgreater{} [1] 119.0  58.6  36.9  28.2}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df }\OtherTok{\textless{}{-}} \FunctionTok{predict\_kinematics}\NormalTok{(}
\NormalTok{  m1,}
  \AttributeTok{max\_time =} \DecValTok{6}\NormalTok{,}
  \AttributeTok{frequency =} \DecValTok{100}\NormalTok{,}
  \CommentTok{\# Additional parameters forwarded to get\_air\_resistance}
  \CommentTok{\# Otherwise, defaults are used}
  \AttributeTok{bodymass =} \DecValTok{80}\NormalTok{,}
  \AttributeTok{bodyheight =} \FloatTok{1.85}\NormalTok{,}
  \AttributeTok{barometric\_pressure =} \DecValTok{780}\NormalTok{,}
  \AttributeTok{air\_temperature =} \DecValTok{20}\NormalTok{,}
  \AttributeTok{wind\_velocity =} \FloatTok{0.5}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{center}\includegraphics[width=0.9\linewidth]{paper_files/figure-latex/unnamed-chunk-7-1} \end{center}

@@

Additional information about mixed-effects model performed using the \texttt{nlme} package (\protect\hyperlink{ref-R-nlme}{Pinheiro, Bates, and R-core 2020}) can be obtained using \texttt{summary()} function. S3 method \texttt{coef()} when applied on mixed-model result will return both the fixed and random effects. To create a simple plot of the model, use S3 \texttt{plot()} method.

@@

\begin{center}\includegraphics[width=0.9\linewidth]{paper_files/figure-latex/unnamed-chunk-11-1} \end{center}

@@

\FunctionTok{ggplot}\NormalTok{(}
  \FunctionTok{filter}\NormalTok{(df, distance }\SpecialCharTok{\textless{}} \DecValTok{40}\NormalTok{),}
  \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ distance, }\AttributeTok{y =}\NormalTok{ value, }\AttributeTok{group =}\NormalTok{ athlete, }\AttributeTok{color =}\NormalTok{ athlete)}
\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{(}\DecValTok{8}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{name, }\AttributeTok{scales =} \StringTok{"free\_y"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{geom\_line}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.7}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{ylab}\NormalTok{(}\ConstantTok{NULL}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{xlab}\NormalTok{(}\StringTok{"Distance (m)"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}
    \AttributeTok{legend.position =} \StringTok{"top"}\NormalTok{,}
    \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{()}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.9\linewidth]{paper_files/figure-latex/unnamed-chunk-12-1} \end{center}

@@

\NormalTok{m3 }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_radar}\NormalTok{(}
  \AttributeTok{velocity =}\NormalTok{ sprint\_velocity,}
  \AttributeTok{time =}\NormalTok{ sprint\_time}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m3\_weighted }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_radar}\NormalTok{(}
  \AttributeTok{velocity =}\NormalTok{ sprint\_velocity,}
  \AttributeTok{time =}\NormalTok{ sprint\_time,}
  \AttributeTok{weights =} \DecValTok{1} \SpecialCharTok{/}\NormalTok{ (sprint\_velocity }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}
